To be FAIR: Theory Specification Needs an Update

### 2025-02-23 manuscript.tex
complete draft for final comments

  man,floatsintext]{apa6}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8186}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={To be FAIR: Theory Specification Needs an Update},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

This is a preprint paper, generated from Git Commit \# c8b13da7.

@@

\abstract{%
Many meta‐scientific initiatives have focused on enhancing the rigor of theory testing, yet the systematic specification of theories remains underdeveloped. We present FAIR theory, a new approach that transforms scientific theories into digital, evolving artifacts---developed collaboratively---to bolster the robustness and interconnectedness of scientific knowledge. By representing theories as formal, machine‐readable objects (e.g., mathematical models, computer code, nomological networks, or explicit operational definitions) and tracking their evolution through version control and semantic versioning, FAIR theory provides a structured, cumulative framework for theory development. We propose that each theory is assigned a universally unique identifier and enriched with detailed metadata, ensuring it is Findable, Accessible, Interoperable, and Reusable. This digital infrastructure addresses a critical gap in scholarly practice by introducing the collaborative tools and standards for theory specification that have already revolutionized methodology, data sharing, and publishing. Ultimately, FAIR theory aims to transform how scientific ideas are communicated, evaluated, and iteratively refined, thereby accelerating cumulative knowledge acquisition across disciplines.
}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable {[}REF{]} for both humans and computers.
Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.

@@

The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.

Scholars have been raising concerns about the state of theory in social science for nearly 50 years (Paul E. Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One main concern is that social scientific theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
When theories do not make precise predictions,
they are hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' (\textbf{peirceCollectedPapersCharles1960?}). However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4881). Defining what makes a theory ``best'' requires making auxiliary assumptions. For present purposes, induction and abduction are interchangeable terms.}.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

\begin{figure}
\centering
\includegraphics{empirical_cycle.pdf}
\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
\end{figure}

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected, let alone derived, from theory,
and test results do not routinely reference back to theories, contributing to their improvement or rejection.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Paul E. Meehl, 1978).

@@

The present paper addresses the lack of open science methods for theory development and suggests an improvement of the state of affairs by applying the FAIR principles to scientific theories.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We apply the FAIR principles to digital representations of theory,
introducing a FAIR metadata format to make theories \emph{Findable} via a DOI,
\emph{Accessible} in a machine- and human-readable filetype,
\emph{Interoperable} within the data analysis environment,
and \emph{Reusable} in the practical and legal sense, so that they may be improved over time -- at best, in a participative process.
Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see Table \ref{tab:tabfair}.
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in the social sciences.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

\begin{lltable}

\begin{longtable}{m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}m{.1\linewidth}m{.35\linewidth}m{.35\linewidth}m{.15\linewidth}}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:tabfair}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:tabfair} continued}}\\
\toprule
Criterion & \multicolumn{1}{c}{Original} & \multicolumn{1}{c}{Theory} & \multicolumn{1}{c}{Action}\\
\midrule
\endhead
F1 & (Meta)data are assigned a globally unique and persistent identifier & Theory (meta)data has a global, unique and persistent identifier & Rephrased\\
F2 & Data are described with rich metadata & Theory is described with rich metadata & Rephrased\\
F3 & Metadata clearly and explicitly include the identifier of the data it describes & Metadata clearly and explicitly include identifiers for all the versions of the theory it describes & Rephrased and extended\\
F4 & (Meta)data are registered or indexed in a searchable resource & Theory and its associated metadata are included in a searchable repository & Rephrased, needs work\\
A1 & (Meta)data are retrievable by their identifier using a standardized communications protocol & Theory and its associated metadata are accessible by their identifier using a standardized communications protocol & Rephrased\\
A1.1 & The protocol is open, free, and universally implementable & The protocol is open, free, and universally implementable & Remain the same\\
A1.2 & The protocol allows for an authentication and authorization procedure, where necessary & The protocol allows for an authentication and authorization procedure, where necessary & Remain the same, but less relevant\\
A2 & Metadata are accessible, even when the data are no longer available & Theory metadata are accessible, even when the theory is no longer available & Rephrased, but less relevant\\
I1 & (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge representation & Theory and its associated metadata use a formal, accessible, shared and broadly applicable language to facilitate machine readability and reuse & Rephrased and extended\\
I2 & (Meta)data use vocabularies that follow FAIR principles & (Meta)data use vocabularies that follow FAIR principles & Rephrased\\
I2S.1 & - &  & \\
I2S.2 & - &  & \\
I3 & (Meta)data include qualified references to other (meta)data & (Meta)data includes qualified references to other (meta)data, including previous versions of the theory & Extended\\
I4S & - &  & \\
R1 & (Meta)data are richly described with a plurality of accurate and relevant attributes & Theory and its associated metadata are richly described with a plurality of accurate and relevant attributes & Rephrased\\
R1.1 & (Meta)data are released with a clear and accessible data usage license & Theory (meta)data are released with a clear and accessible license & Rephrased\\
R1.2 & (Meta)data are associated with detailed provenance & Theory (meta)data are associated with detailed provenance & Rephrased\\
R1.3 & (Meta)data meet domain-relevant community standards & Theory (meta)data and documentation meet domain-relevant community standards & Rephrased\\
\bottomrule
\end{longtable}

\end{lltable}

@@

Definitions of theory abound and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
our paper is not aligned with one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021)
The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's hierarchy of ever-more specific ``statements'' a theory might contain (1990):
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values Guest (2024).

The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term ``model'' is not uniquely defined within the literature;
it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' {[}REF Fried{]}.
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
-

The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g.~commonly used analogies) and practices (e.g.~experimental designs) can be an important part of scientific theories.

Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
the method is not contained to any one particular definition.
It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, the nature of the information tracked will likely change.
For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
From this theory, more precise \emph{statistical/mathematical models} could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

Although we are in favor of the advancement towards formal theories,
this might not always be desirable or feasible.
Fortunately, formalization is not required to make theories FAIR.
To some extent, FAIRness and formalization are orthogonal.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).

@@

We propose FAIR theory as an instantiation of modular publishing (\textbf{kirczModularityNextForm1998?}).
At present, the primary unit of social scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
with adequate metadata that is indexed in standardized repositories (\textbf{vandesompelRethinkingScholarlyCommunication2004?}).
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the \href{}{DataCite Metadata Schema} offers a standard way to document the nature of relationships between information artifacts.
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
The present paper considers the advantages of Git for FAIR theory.
Git enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).

>>>>

  man, noextraspace,floatsintext]{apa6}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8881}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{longtable, setspace}
\raggedbottom
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={To be FAIR: Theory Specification Needs an Update},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

This is a preprint paper, generated from Git Commit \# ecb21ba7.

@@

\abstract{%
Innovations in open science and meta-science have focused on rigorous \emph{theory testing}, yet methods for specifying, sharing, and iteratively improving theories remain underdeveloped. To address these limitations, we introduce \emph{FAIR theory}: A standard for specifying theories as Findable, Accessible, Interoperable, and Reusable information artifacts. FAIR theories are Findable in well-established archives, Accessible in practical terms (open file formats) and in terms of their ability to be understood, Interoperable for specific purposes, e.g., to guide control variable selection, and Reusable such that they can be iteratively improved. This paper adapts the FAIR priciples for theory, reflects on the FAIRness of contemporary theoretical practices in psychology, introduces a workflow for FAIRifying theory, and explores FAIR theories' potential impact in terms of reducing research waste, enabling meta-research on the structure and development of theories, and incorporating theory into reproducible research workflows -- from hypothesis generation to simulation studies. We make use of well-established open science infrastructure, including Git for version control, GitHub for collaboration, and Zenodo for archival and search indexing. By applying the principles and infrastructure that have already revolutionized sharing of data and publications to theory, we establish a sustainable, transparent, and collaborative approach to theory development. FAIR theory equips scholars with a standard for systematically specifying and refining theories, bridging a critical gap in open research practices and supporting the renewed interest in theory development in psychology and beyond. FAIR theory provides a structured, cumulative framework for theory development, increasing efficiency and potentially accelerating the pace of cumulative knowledge acquisition.
}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016a) for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.

@@

The so-called ``replication crisis'' has prompted extensive reforms in psychology and other scientific fields (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in ``self-determination theory'' (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures (Deci, 1971).
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs---autonomy, competence, and relatedness---all still categorized as ``intrinsic'' (Ryan \& Deci, 2000).
The profound difference between these definitions becomes clear when considering the act of changing a child's dirty diaper.
Under the original definition, few would be considered intrinsically motivated, after all, it's not exactly a joy-filled experience
Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one's child.

Scholars have been raising concerns about the state of theory in psychology for nearly 50 years (Meehl, 1978; Robinaugh, Haslbeck, Ryan, Fried, \& Waldorp, 2021).
One main concern is that theories lack precision, or \emph{formalization} (Szollosi \& Donkin, 2021).
When theories do not make precise predictions,
they are hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processeses of theory construction and improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories. Others have used the term ``abduction'', coined by Peirce to describe ``inference to the best explanation''. However, Peirce later acknowledged that abduction ``is not essentially different from induction'' (Peirce, 1960, p. 4881). As defining what makes a theory ``best'' requires auxiliary assumptions, we consider abduction to be a special case of induction. For present purposes, however, the terms are interchangeable.}.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases, see Figure \ref{fig:figecs}, panel (a).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

\begin{figure}
\centering
\includegraphics{ec_visio.pdf}
\caption{\label{fig:figecs}Three implementations of the ``empirical cycle''.}
\end{figure}

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices (Gigerenzer, Krauss, \& Vitouch, 2004),
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected (Van Lissa et al., 2020).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Meehl, 1978).

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are ``data about the data'', they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are \emph{Findable} via a DOI or by searching the repository they are archived in;
\emph{Accessible} in a machine- and human-readable filetype;
\emph{Interoperable} for specific purposed, for example, within the data analysis environment;
and \emph{Reusable} in the practical and legal sense, so that they may be iteratively improved by the author or by others.
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see \href{https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv}{Supplemental Table S1}.
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

Definitions of theory abound and are the subject of extensive scholarly debate.
Given that a pluriformity of definitions are consistent with FAIR theory principles,
the method is not contained to any one particular definition.
Perspectives on scientific theory have been categorized as syntactic, semantic, and pragmatic (Winther, 2021)
The syntactic view describes theories as ``sets of sentences in a given logical domain language'' (Winther, 2021, ch.~2),
acknowledging that each domain (a scientific field, such as psychology or physics) has its own theoretical vocabulary.
We recognize the syntactic view in Meehl's hierarchy of ever-more specific ``statements'' a theory might contain (1990):
statements about the types of entities postulated (i.e., ontology),
statements about causal connections between those entities,
statements about the functional form of those connections,
and statements about their specific numerical values (cf. Frankenhuis, Panchanathan, \& Smaldino, 2023; Guest, 2024).

The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term ``model'' is not uniquely defined within the literature;
it has been described as a ``specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory'' (Fried, 2020).
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
The pragmatic view holds that there might not be one structure or definition of scientific theories, but instead, definitions differ across scientific domains.
It also argues that nonformal aspects (e.g.~commonly used analogies) and practices (e.g.~experimental designs) can be an important part of scientific theories.

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, details may be added, and the nature of the information tracked might even change.
For example, following Meehl, we could envision a theory that starts out with establishing, through observation, an ontology of constructs relevant for a given phenomenon.
After initial exploratory research, the theory might be further specified by making assumptions about how these constructs are causally connected.
Over time, more precise \emph{statistical/mathematical models} could be derived by further assuming a specific functional form for relationships (e.g., linear effects) and error families for the distribution of measured variables (e.g., normal distributions).
This allows for the specification of statistical models, which make just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Going even further, a \emph{generative/computational model} could be specified,
which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time - either to the theory itself, or as references recorded in the theory metadata.
Examples include experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} (Popper, 2002) because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, ``grounded theory'', derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).

@@

We propose FAIR theory as an instantiation of modular publishing (Kircz, 1998).
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
with adequate metadata that is indexed in standardized repositories (Van De Sompel, Payette, Erickson, Lagoze, \& Warner, 2004).
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts (DataCite Metadata Working Group, 2024).
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using \texttt{relationType:\ IsDescribedBy}, while the reverse relationship, documented in the canonical reference paper, is \texttt{relationType:\ Describes}. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as \texttt{relationType:\ IsDerivedFrom}.
If an existing theory is updated, \texttt{relationType:IsNewVersionOf} could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with \texttt{relationType:\ IsVariantFormOf}.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
The present paper considers the advantages of Git for FAIR theory.
Git enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).
